Experience Report: Haskell in Computational Biology

### 2012-03-11 mrfy_experience_report.tex
Fixed NR's botch of the search strategies and their types

MRFy's search works with a sequence of \emph{generations}.
A~generation is a collection of scored \emph{placements}.
We~pass the
search function a \texttt{Scorer}, which is the result of partially
applying \texttt{score} to the model.
\smallverbatiminput{scoredecl}
A~search algorithm is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{strategy}
In this record,
\begin{itemize}
\item
Function \texttt{gen0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may not be any better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} considers the score of the best
placement of the trial generation, along with the history of previous
best scores.
It~also considers the \texttt{Age} of the trial generation, which is
the total number of 
generations tried.
\item
Finally, the \texttt{quit} function decides when to stop the search.
It~too considers the history of scores and the age of the current
generation. 
\end{itemize}
This interface provides ample leeway for experiments.
For~example, random-mutation hill climbing accepts only generations
with better scores than their predecessors,
while simulated annealing may accept a generation with a worse score.
A~genetic algorithm uses \texttt{nextGen} to
combine placements from the previous generation, while
random-mutation
hill climbing simply permutes a placement into a new one.
The interface \texttt{quit} makes it easy
to implement such termination criteria as a fixed number of generations
or a limit of going $k$~generations without achieving more than $E$
improvement in score.

>>>>

MRFy's search works with a sequence of \emph{generations}.
A~generation is a collection of scored \emph{placements}.
We~pass the
search function a \texttt{Scorer}, which is the result of partially
applying \texttt{score} to the model.
\smallverbatiminput{scoredecl}
A~search algorithm is characterized by four functions, which are
bundled into a record called \texttt{SearchStrategy}.
\smallfuzzverbatiminput{2.3pt}{strategy}
In this record,
\begin{itemize}
\item
Function \texttt{gen0} produces the first generation of placements.
\item
Function \texttt{nextGen} produces a new generation of placements from
an existing generation.
Because scoring a placement is expensive, 
and some placements may persist from the old generation to the new
generation, \texttt{nextGen} works with \emph{scored} placements, not
bare placements.
\item
A~new generation produced by \texttt{nextGen} may not be any better
than the generation it replaces.
When it is first produced, the new generation is ``on trial.''
The \texttt{accept} function decides whether to keep a new generation
or to try \texttt{nextGen} again with a different seed.
The \texttt{accept} function considers a list of type
\texttt{[Scored~Age]}.
At~the head of this list is the score of the best
placement of the trial generation, together with the \texttt{Age} of the trial
generation, which is the total number of generations tried.
The remaining elements of the list hold the history of scores and ages for
previously \emph{accepted} generations.
(The list omits generations that didn't survive the \texttt{accept} trial.)
For convenience, the \texttt{accept} function also receives
the \texttt{Age} of the trial generation as an additional parameter.
\item
Finally, the \texttt{quit} function decides when to stop the search.
It~too considers the list of accepted generations and the age of the current
generation. 
\end{itemize}
This interface provides ample leeway for experiments.
For~example, random-mutation hill climbing accepts only generations
with better scores than their predecessors,
while simulated annealing may accept a generation with a worse score.
A~genetic algorithm uses \texttt{nextGen} to
combine placements from the previous generation, while
random-mutation
hill climbing simply permutes a placement into a new one.
The interface \texttt{quit} makes it easy
to implement such termination criteria as a fixed number of generations
or a limit of going $k$~generations without achieving more than $E$
improvement in score.

### 2012-03-11 search.tex
Fixed NR's botch of the search strategies and their types

search :: SearchStrategy -> Scorer -> [Seed]
       -> (Scored Placement, [Score])
search strat score (s:seeds) = runFrom seeds firstGen [] 0
 where
  firstGen = map score $ gen0 strat s
  runFrom :: [Seed] -> [Scored Placement] -> [Score]
          -> Age -> (Scored Placement, [Score])
  runFrom (s1:s2:seeds) oldPop oldHist age =
    let trialPop  = nextGen strat s1 score oldPop
        trialHist = (scoreOf $ minimum newPop) : oldHist
        (newPop, newHist) =
          if accept strat s2 newHist age then
            (trialPop, trialHist)
          else
            (oldPop, oldHist)
    in  if quit strat newHist age then
          (minimum newPop, newHist)
        else
          runFrom seeds newPop newHist (age + 1)

>>>>

search :: SearchStrategy -> Scorer -> [Seed]
       -> (Scored Placement, [Scored Age])
search strat score (s0:seeds) = runFrom seeds firstGen [] 0
 where
  firstGen = map score $ gen0 strat s0
  runFrom :: [Seed] -> [Scored Placement] -> [Scored Age]
          -> Age -> (Scored Placement, [Scored Age])
  runFrom (s1:s2:seeds) oldPop oldHist age =
    let trialPop  = nextGen strat s1 score oldPop
        trialHist = (fmap (const age) $ minimum trialPop)
                  : oldHist
        (newPop, newHist) =
          if accept strat s2 trialHist age then
            (trialPop, trialHist)
          else
            (oldPop, oldHist)
    in  if quit strat newHist age then
          (minimum newPop, newHist)
        else
          runFrom seeds newPop newHist (age + 1)
